---
title: "Predicting E-Commerce Delivery Time Using Linear Regression"
author: "Prepared by: Gibson Gnanasehar - "
format: pdf
editor: visual
---

## Summary

This analysis aims to predict e-commerce delivery time using a linear regression model. The study focuses on the relationship between promised delivery time, regional proximity, and actual delivery time.

## Analysis Code

Installing Packages

```{r}
# Using '#' since these packages are already installed  
#install.packages("tidyverse")
#install.packages("ggplot2")
#install.packages("GGally")
#install.packages("readxl")
```

Loading Libraries

```{r}
library(tidyverse)
library(readxl)
library(GGally)
library(ggplot2)
```

Importing the given data set by read function

```{r}
sales <- read_csv("sales.csv")
```

Viewing the raw data set

```{r}
view(sales)
```

## Data Cleaning

Now removing the missing values in actual_delivery_time column which is our dependable variable for our predictive model.

```{r}
sales_clean = sales |> filter(!is.na(actual_delivery_time))
view(sales_clean)
```

Rounding off actual delivery time value to refine the available data.

```{r}
sales_clean$actual_delivery_time <- round(sales_clean$actual_delivery_time, 2)
```

## Sampling data for Analysis

To ensure efficiency, a random sample of 1000 observation is selected for our analysis

```{r}
id =  #student ID
set.seed(id)
sales_sample <- sales_clean |> slice_sample(n = 1000)


```

Inspecting the final sample

```{r}
view(sales_sample)
glimpse(sales_sample)
```

```{r}
summary(sales_sample)
```

## Choosing Variables

We select *actual_delivery_time* as the dependent variable, with *promise* and *same_region* as independent variables.

## Exploratory Analysis

**Correlation between actual and promised delivery time**

To check correlation of 2 chosen numerical variables for our analysis,

```{r}
cor(sales_sample$actual_delivery_time,sales_sample$promise)
```

The returned correlation value is

```         
0.7208618
```

Which indicates a **strong positive linear relationship,** and it is evident that whenever promised delivery time increases, the actual delivery time also increase.

**Visualizing the relationship**

Also to check this correlation relationship in a plot, we will use scatter plot with regression line as below

```{r}
ggplot(sales_sample, aes(x = promise, y = actual_delivery_time)) +
  geom_point(alpha = 0.5, color = "steelblue") +       
  geom_smooth(method = "lm", se = TRUE, color = "Green") +  
  labs(title = "Relationship Between Promised and Actual Delivery Time",
       x = "Promised Delivery Time (days)",
       y = "Actual Delivery Time (days)") +
  theme_minimal()
```

## Linear Regression Model

Linear regression model fitting as below:-

```{r}
LM <-lm(actual_delivery_time ~ promise + same_region, data = sales_sample)
```

```{r}
summary(LM)
```

R-Square value is **52.2%,** which is the variation in actual delivery time, 

Creating model for the analysis of predicting delivery time

β₀ - Intercept (Actual_delivery time)

β₁ - Regression Coefficient (Promise)

β₂ - Regression Coefficient (Same Region)

ŷ = β₀ + β₁ × promise + β₂ × same_region

The regression equation is **actual_delivery_time = -0.46243 + 0.85531 (Promise) + 0.31589 (Region).**

The above coefficient values are stored in environment.

```{r}
b0 <- LM$coefficients[["(Intercept)"]]
b1 <- LM$coefficients[["promise"]]
b2 <- LM$coefficients[["same_region"]]
```

**Example 1 Prediction:** If an order has promised delivery time of 5 days, and is in same region. Then the predicted delivery time as per the created model will be

**Delivery time = -0.46243+0.85531(5)+0.31589(1) \~= 4.13 days**

```{r}
b0+b1*5+b2*1
```

**Example 2 Prediction:** If an order has promised delivery time of 2 days, and is in different region. Then the predicted delivery time as per the created model will be

**Delivery time = -0.46243+0.85531(2)+0.31589(0) \~= 1.24 days**

```{r}
b0+b1*2+b2*0
```

## Model Application on data set

Create new column 'predicted_delivery' using mutate

```{r}
sales_sample <- sales_sample |>
  mutate(predicted_delivery = round(b0 + b1 * promise + b2 * same_region, 2))
```

## Additional Insight

In logistics, *order_type* is an important variable. However, when assessed using a linear regression model, its statistical significance was low. Despite this, a noticeable difference in delivery time exists between direct (1P) and third-party (3P) orders.

To evaluate this, the data was grouped by *order_type*, and the mean *actual_delivery_time* for each category was calculated as follows:

```{r}
sales_sample |> group_by(order_type) |> 
  summarise(avg_delivery = round(mean(actual_delivery_time), 2))
```

**Visualizing Delivery Time by Order Type**

Creating column chart using group by and summarise function to visualize the above mean.

```{r}
sales_sample |>
  group_by(order_type) |>
  summarise(mean_delivery = mean(actual_delivery_time, na.rm = TRUE)) |>
  ggplot(aes(x = order_type, y = mean_delivery, fill = order_type)) +
  geom_col(width = 0.5, alpha = 0.8) +
  labs(title = "Mean Delivery Time by Order Type",
       x = "Order Type",
       y = "Mean Delivery Time (days)") +
  theme_minimal() +
  theme(legend.position = "none") #ref, Gen AI to remove legend
```

```{r,echo=FALSE,include=FALSE}
ggplot(sales_sample, aes(x = order_type, y = actual_delivery_time, fill = order_type)) +
  geom_boxplot(width = 0.5, alpha = 0.7) +
  stat_summary(fun = mean, geom = "point", shape = 20, size = 3, color = "black") +
  labs(title = "Average Delivery Time by Order Type",
       x = "Order Type",
       y = "Actual Delivery Time (days)") +
  theme_minimal()
```

It is clear that, **1st party direct orders are delivered in \~1 day**, while **3rd party orders take \~3 days**, suggesting operational differences to be addressed.

## Analysis Findings & interpretation

Used linear regression to predict actual e-commerce delivery time based on promised delivery days and region.

Strong positive correlation was found between promised and actual delivery time, and same-region deliveries are generally faster.\

Additional insights reveal that 1st party orders are fulfilled in \~1 day, while 3rd party orders take \~3 days on average.\

These findings help the company to predict delivery time based on model and identify key factors affecting delivery performance, aiding in logistics planning thereby contributes to service improvement.
