---
title: "Predicting Customer's Loan decision by Logistic Regression"
author: "Gibson Gnanasehar - Student ID: "
format: 
  pdf: 
    geometry:
      - top=14mm
      - left=15mm
      - right=15mm
      - bottom=12mm
      - heightrounded
    colorlinks: true
    papersize: a4
    fontsize: 11pt
    mainfont: Times New Roman
    sansfont: Times New Roman
    align: justify
    include-in-header:
      text: |
        \usepackage{titling}
        \usepackage{tikz}
        \usetikzlibrary{calc}
        \usepackage{tikzpagenodes}
        \usepackage{atbegshi}
        \usepackage{setspace}
        \usepackage{fancyhdr} % enable custom headers/footers
        \usepackage{wrapfig} % ADDED for wrapping
        \usepackage{float} % ADDED for fixed placement
        \usepackage{caption}
        \usepackage{subcaption} % ADDED for side-by-side layout
        \onehalfspacing
        \pagestyle{fancy}
        \fancyhf{} % clear default header/footer
        \fancyhead[R]{\thepage} % page number on top right
        \renewcommand{\headrulewidth}{0pt} % no header line
        % Adjusting title position and font size
        \pretitle{%
          \begin{center}
          \vspace*{-2cm}
          \fontsize{14pt}{16pt}\selectfont\bfseries
        }
        \posttitle{\par\end{center}\vspace*{-0.75cm}}
        % Adjusting author position and font size
        \preauthor{%
          \begin{center}
          \fontsize{11pt}{13pt}\selectfont
        }
        \postauthor{\end{center}\vspace*{-2.25cm}}
        % Page border
        \AtBeginShipout{\AtBeginShipoutAddToBox{%
          \begin{tikzpicture}[remember picture,overlay]
            \draw[line width=0.5pt]
              ($(current page.north west)+(1cm,-1cm)$) rectangle
              ($(current page.south east)+(-1cm,1cm)$);
          \end{tikzpicture}%
        }}
editor:
  markdown:
    wrap: 72
editor_options:
  chunk_output_type: console
execute:
  warning: false
  message: false
  echo: false
---

```{r, include=FALSE}
# install.packages('tidyverse')
# install.packages('dplyr')
# install.packages('ggplot2')
# install.packages('tibble')
# install.packages('stringr')
# install.packages('tinytex')
# install.packages('broom')
# install.packages('kableExtra')
# install.packages('GGally')
# install.packages('cowplot')
# install.packages('knitr')
# tinytex::tlmgr_install("wrapfig")
# install.packages("webshot2")
# webshot2::install_phantomjs()
```

```{r setup, include=FALSE}
# Load required libraries
library(tidymodels)
library(ggplot2)
library(readr)
library(skimr)
library(kableExtra)
library(tidyverse)
library(dplyr)
library(tibble)
library(stringr)
library(tinytex)
library(broom)
library(GGally)
library(cowplot)
library(knitr)
library(webshot2)
```

## Introduction

This report studies the factors influencing personal loan acceptance
using a logistic regression model. The model is performed on a random
sample of 500 observations drawn from the *bankloan.csv* dataset. The
binary outcome variable is *personal_loan* (Yes/No), and key predictors
include *Income*, *Family*, *Education*, *CD Account*, and *CreditCard*
usage. The objective is to identify customer attributes that
significantly influence loan uptake, enabling data-driven targeting and
credit risk assessment in banking.

### Data Preparation

A fixed random seed ensured reproducibility during sampling. Missing
data were excluded, and categorical variables such as *Education*, *CD
Account*, and *CreditCard* were converted to factor type. The target
variable was relabeled for interpretability. Only predictors relevant to
the loan decision were retained.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
id <- 
set.seed(id)
all_data <- read_csv("bankloan.csv")
data_sample <- all_data |> slice_sample(n = 500)

# Clean and select variables
clean_data <- data_sample |>
  select(personal_loan, Income, Family, Education, `CD Account`, CreditCard) |>
  na.omit() |>
  mutate(
    personal_loan = factor(personal_loan, levels = c(0, 1), labels = c("No", "Yes")),
    `CD Account` = factor(`CD Account`),
    CreditCard = factor(CreditCard),
    Education = factor(Education),
    Family = as.integer(Family)
  )
```

## Data Visualization

*Figure 1* demonstrates that customers who accepted personal loans
exhibited significantly higher income levels compared to those who did
not, indicating income as a key predictive factor.

```{r, echo=FALSE}
fig1 <- ggplot(clean_data, aes(x = personal_loan, y = Income)) +
  geom_boxplot(fill = "lightblue") +
  labs(x = "Personal Loan", y = "Income") +
  theme_minimal(base_size = 8) +
  theme(axis.title = element_text(size = 11), axis.text = element_text(size = 10)
  )
ggsave("fig1.pdf", fig1, width = 5, height = 3)
```

```{r, echo=FALSE}
fig2 <- ggplot(clean_data, aes(x = `CD Account`, fill = personal_loan)) +
  geom_bar(position = "fill") +
  labs(x = "CD Account", y = "Proportion") +
  theme_minimal(base_size = 8) +
  theme(axis.title = element_text(size = 11), axis.text = element_text(size = 10)
  )

ggsave("fig2.pdf", fig2, width = 5, height = 3)

```

A larger proportion of customers with CD accounts accepted loans as
illustrated in *Figure 2*, suggesting a strong relationship between
existing product engagement and loan uptake.

```{=latex}
\noindent
\makebox[0.45\textwidth][l]{\textbf{Figure 1: Income by Loan Status}}%
\hfill
\makebox[0.45\textwidth][r]{\textbf{Figure 2: CD Account Ownership by Loan Status}}

\begin{figure}[H]
\centering
\begin{subfigure}[b]{0.45\textwidth}
\includegraphics[width=\textwidth]{fig1.pdf}
\end{subfigure}\hfill
\begin{subfigure}[b]{0.45\textwidth}
\includegraphics[width=\textwidth]{fig2.pdf}
\end{subfigure}
\end{figure}
```

**Table 1: Mean Income by Loan Status and CD Account**

```{r}
summary_table <- clean_data |> 
  group_by(personal_loan, `CD Account`) |> 
  summarise(Mean_Income = mean(Income), .groups = 'drop')

table1 <- summary_table |>
  kable(format = "latex", booktabs = TRUE, align = "l") |>  # "l" = left alignment
  kable_styling(font_size = 10, position = "left", latex_options = "hold_position")

table1
```

*Table 1* confirmed that mean income among acceptors was nearly double
that of non-acceptors, regardless of CD account ownership, highlighting
both variables as key influencers.

```{r}
# Fit the logistic regression model globally (outside layout blocks)
set.seed(123)
data_split <- initial_split(clean_data, prop = 0.7)
train_data <- training(data_split)
test_data <- testing(data_split)

model_rec <- recipe(personal_loan ~ Income + Family + Education + `CD Account` + CreditCard, data = train_data)
log_reg <- logistic_reg() |> set_engine("glm")
model_wflow <- workflow() |> add_model(log_reg) |> add_recipe(model_rec)
model_fit <- model_wflow |> fit(data = train_data)

```

**Logistic Regression Model**

A logistic regression model was considered due to the binary nature of
the outcome. The model included five predictors: *Income*, *Family*,
*Education*, *CD Account*, and *CreditCard*.

*Table 2* presents the regression coefficients, standard errors, test
statistics, and p-values for each predictor.

**Table 2: Logistic Regression Coefficients**

```{r}
# Table 2: Logistic Regression Coefficients
coeff_table <- model_fit |> 
  extract_fit_parsnip() |> 
  tidy()

coeff_table |>
  kable(format = "latex", booktabs = TRUE, align = "l") |>  # left-align all columns
  kable_styling(font_size = 10, position = "left", latex_options = "hold_position")
```

As shown in Table 2, *Income* (β = 0.0605, *p* \< .001) and *Family*
size (β = 0.6024, *p* = .0029) had significant positive effects,
indicating higher chances of loan acceptance among wealthier individuals
and larger families. Education levels also mattered: those with a higher
education were significantly more likely to accept loans (*p* \< .001).
*CD Account* ownership had the strongest effect (β = 3.22, *p* \< .001),
reflecting deeper financial engagement. *CreditCard* ownership showed a
negative but statistically insignificant effect on loan acceptance.
However, from a behavioral perspective, customers who actively use
credit cards may exhibit a borrowing nature and thus be more inclined
towards loans. The model's intercept (β = –10.41) represents the low
probability when all predictors are at the base level.

## Model Performance

The logistic regression model was applied on the 30% test set to assess
how well it predicts loan acceptance.

```{r, echo=FALSE}
predictions <- augment(model_fit, test_data)
```

**Table 3: Confusion Matrix**

```{r}
# Confusion matrix and its formatted version
confusion_tbl <- conf_mat(predictions, truth = personal_loan, estimate = .pred_class)
```

```{r}
conf_mat_df <- as.data.frame.matrix(confusion_tbl$table)

knitr::kable(
  conf_mat_df,
  format = "latex",
  booktabs = TRUE,
  align = rep("l", ncol(conf_mat_df) + 1)  # +1 for row names if included
) |>
  kableExtra::kable_styling(
    position = "left",                      # Aligns full table block to left
    latex_options = c("hold_position", "scale_down"),
    font_size = 10
  )
```

The confusion matrix results shown in *Table 3* highlight strong
classification performance. The model correctly identified 50 loan
acceptors (true positives) and 85 non-acceptors (true negatives).

The model correctly classified 90% of all cases, indicating high overall
predictive
performance. Sensitivity (Recall) = 0.847: About 84.7% of actual loan
acceptors
were correctly identified. Specificity = 0.934: Around 93.4% of
non-acceptors were accurately predicted, indicating very low false
positives (ref: *Table 4*).

**Table 4: Model Performance Metrics**

```{r, echo=FALSE}
# Calculate metrics in separate named objects
accuracy_val <- predictions |> yardstick::accuracy(truth = personal_loan, estimate = .pred_class, event_level = "second")
sensitivity_val <- predictions |> yardstick::sensitivity(truth = personal_loan, estimate = .pred_class, event_level = "second")
specificity_val <- predictions |> yardstick::specificity(truth = personal_loan, estimate = .pred_class, event_level = "second")
auc_val <- predictions |> roc_auc(truth = personal_loan, .pred_Yes, event_level = "second")

# Combine and rename metrics
performance_table <- bind_rows(
  accuracy_val |> mutate(.metric = "accuracy"),
  sensitivity_val |> mutate(.metric = "sensitivity"),
  specificity_val |> mutate(.metric = "specificity"),
  auc_val |> mutate(.metric = "auc")
)
performance_table|> knitr::kable(format = "latex", booktabs = TRUE, align = "l") |>  # Left-align all columns
  kableExtra::kable_styling(
    position = "left",                # Aligns entire table to the left
    latex_options = "hold_position", # Prevents floating
    font_size = 10                    # Sets font size
  )
```

Together, these results confirm that the logistic regression model is
highly effective and reliable for predicting personal loan acceptance in
this dataset.

The ROC curve, as shown in *Figure 3,* indicates excellent separation
between the two outcome classes.

**Figure 3: ROC Curve for Logistic Regression Model**

```{r fig-roc-curve, fig.width=3, fig.height=2, fig.align='center', echo=FALSE, fig.cap=""}
roc_curve(predictions, truth = personal_loan, .pred_Yes, event_level = "second") |> 
  autoplot()
```

## Conclusion

The logistic regression model displays strong predictive performance
(90% accuracy, AUC = 0.96). Key predictors such as *Income, Education,
Family* size, and *CD Account* ownership were significantly associated
with higher loan acceptance likelihood. Based on this model banks can
improve targeting by focusing on financially stable, larger families,
well-educated customers already engaged with their products, thereby
enhancing marketing efficiency and minimizing credit risk.
